---
title: "Assignment 3 - Text Mining"
link-citations: true
toc: false
format: 
  html:
    theme: simplex
    embed-resources: true
---

## Due Date

This assignment is due by 11:59pm Eastern Time on Friday, March 27th, 2025, submitted on [Github issues](https://github.com/dmcable/BIOSTAT620W26/issues/12).

## Text Mining

On the course Github, navigate to the page for dataset `pubmed`
(<https://github.com/dmcable/BIOSTAT620W26/tree/main/data/pubmed>) and
download the `pubmed.csv` dataset. This dataset contains 3,241 abstracts from
articles collected via 5 PubMed searches. The search terms are listed in the
second column, `term` and these will serve as the "documents." Your job is to
analyze these abstracts to find interesting insights.

1. (15 points) How many abstracts are associated with each search term? Tokenize
the abstracts and count the number of each token. Do you see anything
interesting? What are the 5 most common tokens for each search term?
```{r}
library(readr)
library(dplyr)
library(tidytext)
library(tidyr)
library(ggplot2)

# Load the dataset
url <- "https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/pubmed/pubmed.csv"
pubmed <- read_csv(url)

# 1. How many abstracts are associated with each search term?
abstract_counts <- pubmed %>%
  count(term, name = "number_of_abstracts") %>%
  arrange(desc(number_of_abstracts))

print("Abstracts per search term:")
print(abstract_counts)

# 2. Tokenize the abstracts and count the number of each token
tokens_all <- pubmed %>%
  unnest_tokens(word, abstract)

# Overall token counts
token_counts <- tokens_all %>%
  count(word, sort = TRUE)

print(head(token_counts, 10))

# 3. What are the 5 most common tokens for each search term? (With stop words)
top_5_tokens_per_term <- tokens_all %>%
  group_by(term) %>%
  count(word, sort = TRUE) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>%
  ungroup()

print("Top 5 tokens per term:")
print(top_5_tokens_per_term)

```
Ans: Abstracts per search term with counts are as above. I found there are some stop words. Top 5 tokens per term are also as above.


2. (15 points) Does removing stop words change what tokens appear as the
most frequent? What are the 5 most common tokens for each search term after
removing stop words?
```{r}
# remove stop words
tokens_cleaned <- pubmed %>%
  unnest_tokens(word, abstract) %>%
  anti_join(stop_words, by = "word")

# Find the 5 most common tokens for each search term
top_5_cleaned <- tokens_cleaned %>%
  group_by(term) %>%
  count(word, sort = TRUE) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>%
  ungroup()

print("Top 5 tokens per term (removing stop words):")
print(top_5_cleaned)
```
Ans: Yes, it changed what tokens appear as the most frequent. The 5 most common tokens for each search term after removing stop words are as above.


3. (15 points) Tokenize the abstracts into bigrams. Find the 10 most common
bigrams and visualize them.

<!--# look at bigrams were neither word is a stop word -->
```{r}
# Load stop words from tidytext
data("stop_words")

# Tokenize abstracts into bigrams
pubmed_bigrams <- pubmed %>%
  unnest_tokens(bigram, abstract, token = "ngrams", n = 2)

# Separate the bigrams, filter out stop words, and recount
bigram_counts <- pubmed_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%

  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>%

  filter(!is.na(word1) & !is.na(word2)) %>%

  count(word1, word2, sort = TRUE) %>%

  unite(bigram, word1, word2, sep = " ")

# Extract the top 10 most common bigrams
top_10_bigrams <- bigram_counts %>%
  slice_max(n, n = 10, with_ties = FALSE)

print(top_10_bigrams)

# Visualize the top 10 bigrams
ggplot(top_10_bigrams, aes(x = reorder(bigram, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Most Common Bigrams in PubMed Abstracts",
    x = "Bigram",
    y = "Frequency Count"
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = 12))
```


4. (15 points) Calculate the TF-IDF value for each word-search term combination
(here you want the search term to be the "document"). What are the 5 tokens from
each search term with the highest TF-IDF value? How are the results different
from the answers you got in Question 1?
```{r}
# Tokenize the text 
tokens_all <- pubmed %>%
  unnest_tokens(word, abstract)

# Count the frequency of each word per search term (document)
word_counts <- tokens_all %>%
  count(term, word, sort = TRUE)

# Calculate TF-IDF
tfidf_data <- word_counts %>%
  bind_tf_idf(word, term, n)

# Find the top 5 tokens with the highest TF-IDF for each search term
top_5_tfidf <- tfidf_data %>%
  group_by(term) %>%
  slice_max(tf_idf, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(term, desc(tf_idf))

print("Top 5 tokens per term with the highest TF-IDF value:")
print(top_5_tfidf %>% select(term, word, tf_idf, n))
```
Ans: The 5 tokens from each search term with the highest TF-IDF value are as above. The results from question 4 did not have stop words, whereas answers from question 1 had stop words. Additionally, general medical words (like "patients" or "disease") that might have had high frequencies in question 1 are penalized by TF-IDF because they appear across all the different disease abstracts.


## Sentiment Analysis

5. (20 points) Perform a sentiment analysis using the NRC lexicon. What is the
most common sentiment for each search term? What if you remove `"positive"` and
`"negative"` from the list?
```{r}
# install.packages("textdata")
library(textdata)

tokens <- pubmed %>%
  unnest_tokens(word, abstract)

# Get the NRC sentiment lexicon
nrc_lexicon <- get_sentiments("nrc")

# Join tokens with the NRC lexicon and count sentiments per search term
sentiment_counts <- tokens %>%
  inner_join(nrc_lexicon, by = "word", relationship = "many-to-many") %>%
  count(term, sentiment, sort = TRUE)

# What is the most common sentiment for each search term?
top_sentiment_overall <- sentiment_counts %>%
  group_by(term) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup()

print("Most common sentiment for each search term:")
print(top_sentiment_overall)

# What if you remove "positive" and "negative"?
top_sentiment_specific <- sentiment_counts %>%
  filter(!sentiment %in% c("positive", "negative")) %>%
  group_by(term) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup()

print("Most common sentiment removing positive/negative:")
print(top_sentiment_specific)
```
Ans: The most common sentiments for each search term are as above.


6. (20 points) Now perform a sentiment analysis using the AFINN lexicon to get
an average positivity score for each abstract (hint: you may want to create a
variable that indexes, or counts, the abstracts). Create a visualization that
shows these scores grouped by search term. Are any search terms noticeably
different from the others?
```{r}
afinn_lexicon <- get_sentiments("afinn")

abstract_sentiment <- pubmed %>%
  mutate(abstract_id = row_number()) %>%
  unnest_tokens(word, abstract) %>%
  left_join(afinn_lexicon, by = "word") %>%
  mutate(value = replace_na(value, 0)) %>%
  group_by(term, abstract_id) %>%
  summarize(mean_score = mean(value), .groups = "drop")

ggplot(abstract_sentiment, aes(x = reorder(term, mean_score, FUN = median), y = mean_score, fill = term)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 1) +
  coord_flip() + 
  labs(
    title = "Average AFINN sentiment score per abstract",
    x = "Search Term",
    y = "Average positivity score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
Ans: No search term is noticeably different, but cystic fibrosis is more positive and prostate cancermore negative in this data.