---
title: "Lab 8 - Efficiency and Parallelization"
link-citations: true
toc: false
---

# Learning goals

In this lab, you are expected to learn/put in practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the `parallel` package.
```{r}
# install.packages("parallel")
library("parallel")
```


## Problem 1: Vectorization

The following functions can be written to be more efficient without using
parallelization. Write a faster version of each function and show that (1) the
outputs are the same as the slow version, and (2) your version is faster.

1. This function generates an `n x k` dataset with all its entries drawn from a
Poission distribution with mean `lambda`.

```{r p2-fun1, eval = FALSE}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n){
    x <- rbind(x, rpois(k, lambda))    
  }
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
# Generate all n*k draws at once, then format as an n x k matrix
  matrix(rpois(n * k, lambda), nrow = n, ncol = k, byrow = TRUE)
}

# Verify Dimensions and Distributions
set.seed(42)
mat1 <- fun1()

set.seed(42)
mat2 <- fun1alt()

# Show dimensions
dim(mat1)
dim(mat2)
```

Show that `fun1alt` generates a matrix with the same dimensions as `fun1` and
that the values inside the two matrices follow similar distributions. Then check, using `microbenchmark::microbenchmark`,
the speed of the two functions:

```{r check, eval=FALSE}
# Benchmarking
# install.packages("microbenchmark")
library(microbenchmark)

# Check the speed difference
microbenchmark(
  original_loop = fun1(),
  vectorized_mat = fun1alt(),
  times = 100
)
```

2.  This function finds the maximum value of each column of a matrix.

```{r}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

# install.packages("matrixStats")
library(matrixStats)
fun2alt <- function(x) {
  colMaxs(x)
}
```

Show that both functions return the same output for a given input matrix, `x`.
Then check the speed of the two functions, as you did above.
```{r}
# Verify Outputs are the Same
all.equal(fun2(x), fun2alt(x))

# Benchmarking
library(microbenchmark)
microbenchmark(
  original_apply = fun2(x),
  matrixStats = fun2alt(x),
  times = 100
)
```


## Problem 2: Parallelization

We will now turn our attention to the statistical concept of
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)). Among
its many uses, non-parametric bootstrapping allows us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.
Don't worry if these concepts are unfamiliar, we only care about the computation
methods in this lab, not the statistics.

The main assumption is that we can approximate the results of many repeated
experiments by resampling observations from our original dataset, which reflects
the population.

1. This function implements a serial version of the bootstrap. Edit this
function to parallelize the `lapply` loop, using whichever method you prefer.
Rather than specifying the number of cores to use, use the number given by the
`ncpus` argument, so that we can test it with different numbers of cores later.

```{r p3-boot-fun, eval = FALSE}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n * R, TRUE), nrow = n, ncol = R)
  
  if (ncpus > 1L) {
    # 1. Create a background cluster with the specified number of cores
    cl <- makeCluster(ncpus)
    
    # 2. Export necessary variables and functions to the cluster workers
    clusterExport(cl, varlist = c("dat", "idx", "stat"), envir = environment())
    
    # 3. Run the parallelized loop
    ans <- parLapply(cl, seq_len(R), function(i) {
      stat(dat[idx[, i], , drop = FALSE])
    })
    
    # 4. Shut down the cluster to free up memory
    stopCluster(cl) 
    
  } else {
    # If ncpus is 1, just run the standard, serial lapply
    ans <- lapply(seq_len(R), function(i) {
      stat(dat[idx[, i], , drop = FALSE])
    })
  }
  
  # Converting the list into a matrix
  ans <- do.call(rbind, ans)
  return(ans)
}
```


2. Once you have a version of the `my_boot()` function that runs on multiple
cores, check that it provides accurate results by comparing it to a parametric
model:

```{r p3-test-boot, eval = FALSE}
# Bootstrap of an OLS
my_stat <- function(d) coef(lm(y ~ x, data=d))

# DATA SIM
set.seed(1)
n <- 500; R <- 1e4

x <- cbind(rnorm(n)); y <- x*5 + rnorm(n)

# Checking if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)

# Compare the results
cat("--- Parallel Bootstrap Results ---\n")
print(t(apply(ans1, 2, quantile, c(.025, .975))))

cat("\n--- Standard lm() Confint Results ---\n")
print(ans0)

# You should get something like this
t(apply(ans1, 2, quantile, c(.025,.975)))
##                   2.5%      97.5%
## (Intercept) -0.1372435 0.05074397
## x            4.8680977 5.04539763
ans0
##                  2.5 %     97.5 %
## (Intercept) -0.1379033 0.04797344
## x            4.8650100 5.04883353
```

3. Check whether your version actually goes faster when it's run on multiple
cores (since this might take a little while to run, we'll use `system.time` and
just run each version once, rather than `microbenchmark`, which would run each
version 100 times, by default):

```{r benchmark-problem3, eval = FALSE}
# Find out how many cores the Windows machine has, save one for the OS
num_cores <- max(1L, detectCores() - 1L) 

cat("Timing Single Core (ncpus = 1)...\n")
time_serial <- system.time({
  my_boot(dat = data.frame(x, y), stat = my_stat, R = R, ncpus = 1L)
})
print(time_serial)

cat("\nTiming Multiple Cores (ncpus =", num_cores, ")...\n")
time_parallel <- system.time({
  my_boot(dat = data.frame(x, y), stat = my_stat, R = R, ncpus = num_cores)
})
print(time_parallel)

```

## Problem 3: Short answer

1. Explain in your own words how you can decide if a problem is fit for parallelization? 

Ans: I think two points to consider, 1) task dependence, If I have computations A and B, A and B can be done independently and have same results, then parallelization can be considered; 2) cost from coputational loading, if the heavy computational loading is costly, then we can consider parallelization to save the budget.

2. We have covered vectorization and parallelization in this lab. Explain in your own words how you can think about the relationship between vectorization and paralleization? How do we know whether we should use one or the other or both?

Ans: Vectorization happens in a single CPU core, while paralleization happens in multiple CPU cores. We may consider vectorization first because it only requires a single CPU core. But if you still take much time in computation, then you may consider paralleization.

